---
title: ReadMe.rmd
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code is designed to download and take the files from the UCI HAR Dataset, extract the mean and the standard deviation for some of the normalized smartphone measures, and then create a tidy data table containing the mean of those collected measures.  The collected measures are the mean and the standard devation of the measurments of 30 subjects for 6 activities during the course of an experiment.

The code is commented and begins by creating a data directory.  
It then downloads and unzips the data set.
After it builds out the test and train data and then collects the relevant columns by selecting (using the grepl function) those that have the text "mean()" or "std()" with matching pairs by focusing on those columns whose names end with these measures.

It then appends the subject(called "Participant") and activity numbers.  Using a merge, the activity names are introduced to replace the activity numbers and columns are reorganized.  Since the column vector names and the acivity descriptions are technically descriptive, I maintained them from the original documents.

Finally, I pivoted the data to create a long form and tidy dataset where each row has only one measure and the first 3 columns are unique combinations throughout the dataset and do not duplicate information, other than referring to different measurements.

Finally, I calculated the mean of the means and the standard deviations of the measures collected, thereby creating unique combinations of PArticipant, Activity and Measure for each MeasureMEan, making the dataset tidy in the long form.


```{r cars}
# runanalyis.R
# created to make tidy table of means and standard deviations out of
# test and train dataset provded for the exercise.  Source Dataset is referenced
# README.md and Codebook.md

# check if data folder exists, if not, creates it

if (!dir.exists("./data")){ 
        dir.create("./data")
}

# downlads and unzips the source files in the data directory 
setwd("./data")
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
temp <- tempfile()
download.file(fileUrl,temp)
unzip(temp)
unlink(temp)
setwd("..")


# reads data from test and train files for measures(x), activities(y) and subjects (S)
xtest <- read.table("./data/UCI HAR Dataset/test/X_test.txt")
xtrain<- read.table("./data/UCI HAR Dataset/train/X_train.txt")
ytest <- read.table("./data/UCI HAR Dataset/test/y_test.txt")
ytrain <-read.table("./data/UCI HAR Dataset/train/y_train.txt")
stest <- read.table("./data/UCI HAR Dataset/test/Subject_test.txt")
strain <- read.table("./data/UCI HAR Dataset/train/Subject_train.txt")
features <- read.table("./data/UCI HAR Dataset/features.txt")
activity_labels <- read.table("./data/UCI HAR Dataset/activity_labels.txt")



#merge test and train data for the measures and label the columns
datax <- rbind(xtest,xtrain)
names(datax) <-features[,2]

#keep the mean and std() - I think they come in pairs, so I only kept the activities whose 
#means had a corresponding std() after looking through the list of 
#activities and ignored the meanFreq() etc
datax <- datax[which(grepl("mean[()]",names(datax))|grepl("std[()]",names(datax)))]

#append information on subject and activity
datay <- rbind(ytest,ytrain)
datas <- rbind(stest,strain)
dataset <- cbind(datas,datay,datax)

# read features and activity_labels

names(dataset)[1] <- "Participant" 
names(dataset)[2] <- "Activity"

# name the activities and remove the activities numbers
mdataset <- merge(dataset, activity_labels, by.x="Activity", by.y = "V1") 
mdataset <-subset(mdataset, select = -Activity)


#reorganize the columns so that the activities are closer 
# to the participants and give the activity column a proper name
mdataset <- mdataset[,c(1,68,2:67)]
mdataset
names(mdataset)[2] <- "Activity"


# we have a tidy wide form in mdataset with each row and column 
# intersecting at one  result with no duplicates.  
# use tidyr and dplyr libraries to create a tidy long form dataset
# and calculate the means of the means and the means of the 
# standard deviations.
library(dplyr)
library(tidyr)

calcdataset <- as_tibble(mdataset)

# create a tidy long dataset with a new column Measures to accommodate
# all the readings
calcdataset_long <- pivot_longer(calcdataset,
                 names(calcdataset[,3:68]),names_to =  "Measures")


x<-     calcdataset_long %>%
        group_by(Participant,Activity,Measures) %>%
        summarize(across(everything(),mean),.groups="keep") %>%
        print

names(x)[4] <- "MeasuresMean"
# write the result of the tidy means and averages to the output

write.table(x,"tidyresult.txt",row.name=FALSE)
        
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
